RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing

### 2013-09-11 doc/bioinformatics2013/RVD2.tex
Merged RVD2.tex

%%%%%%%%%%%%%%%%%%
% Chi^2 Test
%%%%%%%%%%%%%%%%%%
\subsection{$\chi^2$ test for non-uniform base distribution}

We assume that at one position in one experimental replicate, the non-reference reads follows multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.

The non-reference reads can be because of either mutation or various errors including sampling error, inter-replicated error, $etc$. In theory, the non-reference read counts caused by non-oriented errors should be uniformly distributed among three non-reference categories, namely $p_1=p_2=p_3=1/3$. However, the distribution of non-reference read counts among three non-reference categories caused by mutation theoretically is biased.

The null hypothesis can be denoted as
\begin{align}
 H_0: p=p_0
\end{align}
where
\begin{align} 
 p_0=(p_1,p_2,p_3), p_1=p_2=p_3=1/3
\end{align}

The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984). Considering observed frequencies $\{r_i; i=1,...,k\}$ and expected frequencies $\{E_i; i=1,...,k\}$, by definition the power-divergence family of statistcs indexed by real parameter $\lambda$ is 
\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R
\end{equation}

Person's $X^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family. Cressie and Read (1984) also proposed that when the $\lambda =2/3$, the new power-divergence statistics outcompete the traditional power-divergence statistics.

Particulaly in our case, Let $(r_1,r_2,r_3)$ denotes the non-reference base frequencies and $\sum_{i=1}^3 r_i = r$. The expected frquencies are $(rp_1, rp_2, rp_3)$. Then the power-divergence statistics can be written as

\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^3 r_i \left[\left(\frac{r_i}{rp_i}\right)^\lambda-1\right];\lambda \in R
\end{equation}

We use $\lambda =2/3$ for the goodness of fit testing.

@@

\section{Results}
Figure 1 Overall of the process
\subsection{Simulation Results}
\subsection{Comparison Results on Synthetic DNA}
Figure 3 ROC comparison with other methods: varscan/samtools/rvd
\subsection{Empirical Results on Multiple Sclerosis Data}
We tested our method on sequence data from clinical Multiple Sclerosis samples. We expect to see that

Table: Table of Variants
\subsection{Performance with read depth}
Figure 4 ROC by read depth
picard to thin segment data, showing performance as read depth decreasing

>>>>

%%%%%%%%%%%%%%%%%%

% Chi^2 Test

%%%%%%%%%%%%%%%%%%
\subsection{$\chi^2$ test for non-uniform base distribution}

We assume that at one position in one experimental replicate, the non-reference read count follows a multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.

The non-reference reads can be because of either mutation or various errors including sampling error or inter-replicated error. We assume, non-reference read counts caused by these random mechanisms should be uniformly distributed among three non-reference categories, giving $p_1=p_2=p_3=1/3$. In contrast, the distribution of non-reference read counts among three non-reference categories caused by mutation would be biased to one base.

We use a $\chi^2$ goodness-of-fit test to distinguish these two possible scenarios. The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$. Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistics, the log likelihood ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$. The test statistic is
\begin{equation}
 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
\end{equation}
where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis. Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.

For each position, there are $N$ power-divergence statistics for each experimental replicate. We use Fisher's combined probability test to combine the p-values from each replicates into a single one for each position. 
\begin{equation}
	X^2 = -2 \sum_{i=1}^N \ln(p_i)
\end{equation}

To control for Bonferroni correction is applied to maintaining the familywise error rate (FWER). The significance level is chosen as 0.05.

@@

\section{Results}

\subsection{Overall of the process}
Figure 1

\subsection{Performance with read depth}\
Figure 2 ROC plot showing performance with different read depth

\begin{figure}[h]
\begin{center}
\includegraphics[width=160mm]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
\label{fig:ROC_without_chi2}
\end{center}
\end{figure}

\begin{figure}[h]
\begin{center}
\includegraphics[width=160mm]{pdf_figs/MuBarPlot.pdf}
\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
\label{fig:ROC_without_chi2}
\end{center}
\end{figure}

We tested the mutation detection power of our graphical model on synthetic dataset using a two-stage test scheme with dilution rates $0.1\%$, $0.3\%$, $1.0\%$, $10.0\%$ and read depth varying from $10^2$ to $10^5$.Different read depth dataset was achieve by thining dataset with read depth up to $10^6$ using Picard. $\chi^2$ test, the second stage test can be optionally included based on average read depth.

Excluding $\chi^2$ test, the ROC curves of our graphical model using Bayesian Hypothesis Testing with different threshold $\tau$ is shown in Figure??. The model had high detection power when the read depth was at $10^5$ level. At dilution rate at $0.1\%$, the model achieved performance with sensitivity at 99.74\% and specificity at 92.86\%. We missed one mutation at position 205 and had one false positive at position 8. The threshold was chosen at 0.022\% which achieved the optimal sensitivity and specificity in the ROC curve. As expected, the detection power of the graphical model decreased when the read depth was reduced.

\begin{figure}[h] %needs to go to supplementary document.
\begin{center}
\includegraphics[width=160mm]{pdf_figs/ROC_subplots_with_chi2.pdf}
\caption{ROC curve varying read depth showing method performance with Bayesian Hypothesis Testing and Chi-Square Testing}
\label{fig:R}
\end{center}
\end{figure}

Figure ?? shows the ROC curves when both Bayesian Hypothesis Testing and $\chi^2$ testing are carried out for variant calling. Comparing with Figure ??, it can be seen that when read depth is low (from $10$ to $10^3$), the false positive rate is severely limited to zero, which means we are very confident in our variant callings. However, the low false positve rate can be achieved at a sacrifice of low true postive rate especially when the dilution rate and read depth are low. The $\chi^2$ test is a stringent test when the read depth is low.

However, if the read depth is high enough (up to $10^4$ or higher), the $\chi^2$ test can limit the false positive rate in a slight degree but no longer at a sacrifice of True Positive Rate. Actually the performance is not much influenced by $\chi_2$ testing with high read depth.

\subsection{Comparison Results on Synthetic DNA}\

Figure 3 ROC comparison with other methods: varscan/samtools/rvd

\subsection{Empirical Results on Multiple Sclerosis Data}\
We tested our method on sequence data from clinical Multiple Sclerosis samples. We expect to see that

Table: Table of Variants
